### 2019-05-30 treestats_paper.tex
Remove trailing whitespace.
Minor edits.

\plr{Review of tree shape stuff in phylogenetics}
a good review of theoretical properties \citet{semple2003phylogenetics}.
\jk{This formulation of a $k$-vector associated with each node is very
reminiscent of classical phlogenetics algorithms. Eg the Sankoff algorithm
for counting the number of mutations needed to explain a set of genotypes
under parsimony and with arbitrary weighting matrices. See
\citep[pg.13]{felsenstein2004inferring}. We should emphasise this connection.}

@@

Suppose we have sequenced a linear genome of $L$ bases from $n$ (homologous) sampled chromosomes,
and wish to compute a statistic of the allele frequency spectrum.
In other words, if the frequency of allele $a$ at a site $i$ is $p_i(a)$,
and the allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $h(p)$,
then for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple average over sites, and monomorphic sites do not contribute,
we can compute this, summing over the $L$ loci, as
$$\sum_p h(p) f(p) = (1/L) \sum_{i=1}^L \sum_a f(p_i(a)).$$
(Note that since the ancestral allele is included, this is a function of the 
\emph{folded} frequency spectrum, so is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and accounting for those that subsequently mutate).
What we do here is a generalization of this idea.

\jk{Perhaps a diagram would help here also, showing a tree and a $k$-vector on each
node?}

@@

A tree sequence describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome \citep{kelleher2016efficient}.
These trees are highly correlated,
a fact which allows them to be stored and processed very efficiently,
as described in \citet{kelleher2018efficient}.
This is encoded as a sequence of trees, $\treeseq = (T_1, T_2, \ldots, T_{n(T)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = G$,
where $G$ is the length of the genome
and $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the node at the bottom,
so for instance the branch describing a relationship where $u$ is the child of $v$
is associated with $u$.
The length of this branch -- i.e., the difference in birth times of $v$ and $u$ --
we denote $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships
can cause the precise set of samples that lie below that edge to differ in different trees.

%%%%%%%%%%%%%%%%%%
\subsection*{Subtree weights}

@@

\begin{definition}[Subtree weights]
    Given a list of \emph{sample weights} $w$, and a tree $T$,
    define the \emph{subtree weight} $\nw_T(u)$ for each node $u$ to be the sum of all sample weights
    of every node below $u$ in the tree (including $u$ itself):
    \begin{align*}
        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ back to the root of $T$.
\end{definition}

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarising we have access to more than one aspect of each subtree.
Below, in section XXX, we describe how to efficiently
maintain a list of the weights of the subtrees below each node
as we iterate through the tree sequence.

@@

@@

Propagating indicator weights of a subset of samples up the tree
lets us easily count the number of those samples that each node in the tree has inherited.
Averaged across trees,
this tells us what proportion of the sample's genomes were inherited from each node.
Other choices of weights would tell us other information about the nodes.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree below node $u$,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{G} \sum_k L_k f(\nw_{T_k}(u)) ,
\end{align}
where the sum is over trees, and $L_k$ is the \emph{span} of the $k^\text{th}$ tree.
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

@@

\paragraph{What about polarization?}
In the definition above we simply sum over all alleles at each site.
However, sometimes it is useful to distinguish the \emph{ancestral} allele 
(i.e., the allele at the root of the tree) from the remaining derived alleles.
This allows statistics in principle to differentiate ancestral from derived alleles,
information which is available in practice (albeit noisily).
One way to make use of this information is to sum over only derived alleles.
We say a site statistic is \textbf{polarized} if we do this,
and so
\begin{align} \label{eqn:site_unpolarized}
    \sitep(f, \iw)_j
    &=
    \sum_{a \in D_j} f(\aw_j(a)) ,
\end{align}
where $D_j$ denotes the set of all alleles at site $j$ except the allele at the root of the tree.
Note that this may not be quite what is expected,
for instance, if there has been a backmutation to the ancestral allele at some point in the tree,
or if there have been mutations to distinct alleles on different parts of the tree
such that the ancestral allele is no longer present.
However, since these situations depend on multiple mutations occurring at a single site,
they are relatively rare in practice.

@@

This makes it natural to define 
a statistic of tree shape by summing these expected contributions across its branches.
Concretely, the \textbf{branch statistic} of a tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
is defined to be
\begin{align}
    \branch(f, \iw)_T
    &=
    \sum_{k=1}^{n(T)} L_k \left\{ f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right\}  ,
\end{align}
where $t_u$ is the length of the branch above node $u$ in the tree $T$
and $\nw(u)$ is the total weight of the subtree below $u$ (as defined above).
The term $\tiw - \nw_{T_k}(u)$ gives the total weight \emph{not} in the subtree of $T_k$ below $u$,
because the sum of all weights in the tree is always constant,
and so so $\sum_u \nw_T(u) = \tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch above $u$,
and $f(\tiw - \nw_{T_k}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of this branch to $\site(f, \iw)$,
per unit of sequence length that the tree $T$ covers.

In practice it is probably most useful to average branch statistics
over a region of the genome,
which we do by averaging the tree statistics over the region
with weights proportional to the trees' spans.
We write this with similar notation:
\begin{align}
    \branch(f, \iw)_{[a,b)}
    &=
    \frac{1}{b-a} \sum_T \ell_T(a,b) \branch(f, \iw)_T ,
\end{align}
where $\ell_T(a,b)$ is the length of the region in $[a,b)$ that the tree $T$ extends for
(i.e., if $T$ covers the half-open interval $[c,d)$,
then $\ell_T(a,b) = \max(0.0, \min(b,d) - \max(a,c))$).

>>>>

\plr{Review of tree shape stuff in phylogenetics}
a good review of theoretical properties \citet{semple2003phylogenetics}.

@@

Suppose we have sequenced a linear genome of $L$ bases from $n$ (homologous) sampled chromosomes,
and wish to compute a statistic of the allele frequency spectrum.
In other words, if the frequency of allele $a$ at a site $i$ is $p_i(a)$,
and the allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $h(p)$,
then for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple average over sites, and monomorphic sites do not contribute,
we can compute this, summing over the $L$ loci, as
$$\sum_p h(p) f(p) = \frac{1}{L} \sum_{i=1}^L \sum_a f(p_i(a)).$$
(Note that since the ancestral allele is included, this is a function of the
\emph{folded} frequency spectrum, so is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and accounting for those that subsequently mutate).
What we do here is a generalization of this idea.

@@

A tree sequence describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient}.
These trees are highly correlated,
a fact which allows them to be stored and processed very efficiently,
as described in \citet{kelleher2018efficient}.
This is encoded as a sequence of trees, $\treeseq = (T_1, T_2, \ldots,
T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance the branch describing a relationship where $u$ is the child of $v$
is associated with $u$.
The length of this branch -- i.e., the difference in birth times of $v$ and $u$ --
we denote $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships
can cause the precise set of samples that lie below that edge to differ across
trees.

%%%%%%%%%%%%%%%%%%
\subsection*{Defining a statistic}

\jk{I think we're missing an overall overview of how to specify a statistic
and basic definitions of the framework itself. Before getting into the
specifics below, it would be good to give an more abstract idea of
what's happening. Perhaps with reference to a diagram?}

@@

\begin{definition}[Subtree weights]
    Given a list of \emph{sample weights} $w$ and a tree $T$,
    define the \emph{subtree weight} $\nw_T(u)$ for each node $u$ to be the sum of all sample weights
    of every node descending from (and including) $u$ in the tree:
    \begin{align*}
        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ to root in $T$.
\end{definition}

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarising we have access to more than one aspect of each subtree.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees below each node
% as we iterate through the tree sequence.

@@

\begin{definition}[Summary function]
    Given a weight vector $(w_1, \dots, w_k)$, the summary function
    $f: \mathbb{R}^k \rightarrow \mathbb{R}^m$ computes a vector
    $(\sigma_1, \dots, \sigma_m)$ summarising these weights.
\end{definition}

\jk{This definition is pretty weak, but wanted something in there.}

@@

Propagating indicator weights of a subset of samples up the tree
lets us easily count the number of those samples that inherit from each node in
the tree. Averaged across trees,
this tells us what proportion of the sample's genomes were inherited from each node.
Other choices of weights would tell us other information about the nodes.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree below node $u$,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k f(\nw_{T_k}(u)).
\end{align}
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

@@

\paragraph{What about polarization?}
In the definition above we simply sum over all alleles at each site.
However, sometimes it is useful to distinguish the \emph{ancestral} allele
(i.e., the allele at the root of the tree) from the remaining derived alleles.
This allows statistics in principle to differentiate ancestral from derived alleles,
information which is available in practice (albeit noisily).
One way to make use of this information is to sum over only derived alleles.
We say a site statistic is \textbf{polarized} if we do this,
and so
\begin{align} \label{eqn:site_unpolarized}
    \sitep(f, \iw)_j
    &=
    \sum_{a \in D_j} f(\aw_j(a)) ,
\end{align}
where $D_j$ denotes the set of all alleles at site $j$ except the allele at the root of the tree.
Note that this may not be quite what is expected,
for instance, if there has been a back mutation to the ancestral allele at some point in the tree,
or if there have been mutations to distinct alleles on different parts of the tree
such that the ancestral allele is no longer present.
However, since these situations depend on multiple mutations occurring at a single site,
they are relatively rare in practice.

@@

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
Concretely, the \textbf{branch statistic} of a tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
is defined to be
\begin{align}
    \branch(f, \iw)_T
    &=
    \sum_{k=1}^{n(\treeseq)} L_k \left\{ f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right\}  ,
\end{align}
where $t_u$ is the length of the branch above node $u$ in the tree $T$
and $\nw(u)$ is the total weight of the subtree below $u$ (as defined above).
The term $\tiw - \nw_{T_k}(u)$ gives the total weight \emph{not} in the subtree of $T_k$ below $u$,
because the sum of all weights in the tree is always constant,
and so $\sum_u \nw_T(u) = \tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch above $u$,
and $f(\tiw - \nw_{T_k}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of this branch to $\site(f, \iw)$,
per unit of sequence length that the tree $T$ covers.

In practice it is probably most useful to average branch statistics
over a region of the genome,
which we do by averaging the tree statistics over the region
with weights proportional to the trees' spans.
We write this with similar notation:
\begin{align}
    \branch(f, \iw)_{[a,b)}
    &=
    \frac{1}{b-a} \sum_T \ell_T(a,b) \branch(f, \iw)_T ,
\end{align}
where $\ell_T(a,b)$ is the length of the region in $[a,b)$ that the tree $T$ extends for
(i.e., if $T$ covers the half-open interval $[c,d)$,
then $\ell_T(a,b) = \max(0, \min(b,d) - \max(a,c))$).
